Unbiased CVM Algorithm

### 2024-12-20 cvm_unbiased/main.tex
Minor corrections.

The following is a helper lemma:
\begin{lemma}
\label{le:helper}
For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
\[
\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$. Let $A' = A \cap \{a_1,\ldots,a_k\}$.

@@

To get a tail estimate, we use the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A'|) \right \wedge p \geq q) \\
  & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
  & \underset{\textrm{Markov}}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
 & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
 & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$):
\[
  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
The latter can be established by analyzing the function $f(q,\eps) = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

@@

Then
\begin{eqnarray*}
  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wegde p \geq q ) \\
    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp(- n/12)
\end{eqnarray*}

@@

Now we come to the tail bounds. The upper tail bound follows directly from Lemma~\ref{le:helper}:
\begin{lemma}\label{le:upper_tail}
Let $0 < \eps \leq 1$ then:
\[
  L = \prob_{\Omega_l} ( p^{-1} |\chi| \geq (1+\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
\begin{eqnarray*}
  L & =&  \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| \wedge p \geq q) \\
  & \leq & \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| ) \\
  & \underset{\textrm{Le.~\ref{le:helper}}}{\leq}  &  \exp\left(-\frac{n}{12} \eps^2\right)
\end{eqnarray*}
\end{proof}

@@

Substituting $t$ and $h$ and using $|A| = \frac{n}{4q}$, we can see that the lemma is true if
\[
  f(q,\eps) = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
\]
is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
This can be verified by checking that $f(q,0) = 0$ and that the derivative with respect to $\eps$ is non-negative.
\end{proof}
We can now establish the concentration result:
\begin{theorem}
Let $0 < \eps < 1$ and $0 < \delta < 1$ and $n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
\[
  L = \prob_{\Omega_l} \left( | p^{-1} |\chi| - |A| | \geq \eps |A| \right) \leq \delta
\]
\end{theorem}
\begin{proof}
Note that the theorem is trivial if $|A| < n$. If not:
\begin{eqnarray*}
  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right) +
    \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1+\eps) |A| \wedge p \geq q \right) +
    \prob_{\Omega_l} \left( p < q \right) \\
    & \underset{Le.~\ref{le:low_p}-\ref{le:upper_tail}}{\leq} &
    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
    & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3}
\end{eqnarray*}
\end{proof}

\section{Unbiasedness}
Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$). Then we can derive from Lemma~\ref{le:neg_cor_pos} using $h(x) = x$ and Lemma~\ref{le:neg_cor_neg} $h(x) = M+1-x$ that:
\begin{eqnarray*}
  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & \leq 1 & \\
  \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in S)) ] \leq M  
\end{eqnarray*}
which implies $\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
By linearity of expectation we can conclude that
\[
  \expectation_{\Omega_l} [ p^{-1} |\chi| ] = \sum_{s \in A} \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = |A| \textrm{.}
\]

>>>>

Let us start with the upper tail bound:
\begin{lemma}
\label{le:upper_tail}
For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
\[
\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$. Let $A' = A \cap \{a_1,\ldots,a_k\}$.

@@

To get a tail estimate, we use the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
  & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
  & \underset{\textrm{Markov}}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
 & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
 & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$):
\[
  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
The latter can be established by analyzing the function $f(q,\eps) = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

@@

Then
\begin{eqnarray*}
  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wedge p \geq q ) \\
    & \underset{\textrm{Le.~\ref{le:upper_tail}}}{\leq} & \exp(- n/12)
\end{eqnarray*}

@@

@@

Substituting $t$ and $h$ and using $|A| = \frac{n}{4q}$, we can see that the lemma is true if
\[
  f(q,\eps) = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
\]
is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
This can be verified by checking that $f(q,0) = 0$ and that the derivative with respect to $\eps$ is non-negative.
\end{proof}
We can now establish the concentration result:
\begin{theorem}
Let $0 < \eps < 1$ and $0 < \delta < 1$ and $n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
\[
  L = \prob_{\Omega_l} \left( | p^{-1} |\chi| - |A| | \geq \eps |A| \right) \leq \delta
\]
\end{theorem}
\begin{proof}
Note that the theorem is trivial if $|A| < n$. If not:
\begin{eqnarray*}
  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right) +
    \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1+\eps) |A| \wedge p \geq q \right) +
    \prob_{\Omega_l} \left( p < q \right) \\
    & \underset{Le.~\ref{le:upper_tail}-\ref{le:lower_tail}}{\leq} &
    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
    & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3}
\end{eqnarray*}
\end{proof}

\section{Unbiasedness}
Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$). Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that:
\begin{eqnarray*}
  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in S) ] \leq 1 \\
  \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in S)) ] \leq M  
\end{eqnarray*}
which implies $\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
By linearity of expectation we can conclude that
\[
  \expectation_{\Omega_l} [ p^{-1} |\chi| ] = \sum_{s \in A} \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = |A| \textrm{.}
\]
